Data Navigator: An Accessibility-Centered Data Navigation Toolkit

### 2023-09-17 input/paper.tex
change bibliography to generic name
add jsdom
downloaded latest source from overleaf
remove weird 7pt text from html
update code and codebase for VIS paper submission

% ---------------------------------------------------------------------------
% Author guideline and sample document for EG publication using LaTeX2e input
% D.Fellner, v1.15, Dec 14, 2018

\documentclass{egpubl}
\usepackage{eurovis2022}

% --- for  Annual CONFERENCE
% \ConferenceSubmission   % uncomment for Conference submission
% \ConferencePaper        % uncomment for (final) Conference Paper
% \STAR                   % uncomment for STAR contribution
% \Tutorial               % uncomment for Tutorial contribution
% \ShortPresentation      % uncomment for (final) Short Conference Presentation
% \Areas                  % uncomment for Areas contribution
% \MedicalPrize           % uncomment for Medical Prize contribution
% \Education              % uncomment for Education contribution
% \Poster                 % uncomment for Poster contribution
% \DC                     % uncomment for Doctoral Consortium
%
% --- for  CGF Journal
% \JournalSubmission    % uncomment for submission to Computer Graphics Forum
% \JournalPaper         % uncomment for final version of Journal Paper
%
% --- for  CGF Journal: special issue
% \SpecialIssueSubmission    % uncomment for submission to , special issue
\SpecialIssuePaper         % uncomment for final version of Computer Graphics Forum, special issue
%                          % EuroVis, SGP, Rendering, PG
% --- for  EG Workshop Proceedings
% \WsSubmission      % uncomment for submission to EG Workshop
% \WsPaper           % uncomment for final version of EG Workshop contribution
% \WsSubmissionJoint % for joint events, for example ICAT-EGVE
% \WsPaperJoint      % for joint events, for example ICAT-EGVE
% \Expressive        % for SBIM, CAe, NPAR
% \DigitalHeritagePaper
% \PaperL2P          % for events EG only asks for License to Publish

% --- for EuroVis 
% for full papers use \SpecialIssuePaper
% \STAREurovis   % for EuroVis additional material 
% \EuroVisPoster % for EuroVis additional material 
% \EuroVisShort  % for EuroVis additional material

% !! *please* don't change anything above
% !! unless you REALLY know what you are doing
% ------------------------------------------------------------------------
\usepackage[T1]{fontenc}
\usepackage{dfadobe}

\usepackage{cite}  % comment out for biblatex with backend=biber
% \usepackage{tabu}
% ---------------------------
%\biberVersion
\BibtexOrBiblatex
%\usepackage[backend=biber,bibstyle=EG,citestyle=alphabetic,backref=true]{biblatex} 
%\addbibresource{chartability.bib}
% ---------------------------  
\electronicVersion
\PrintedOrElectronic
% for including postscript figures
% mind: package option 'draft' will replace PS figure by a filename within a frame
\ifpdf \usepackage[pdftex]{graphicx} \pdfcompresslevel=9
\else \usepackage[dvips]{graphicx} \fi

\usepackage{egweblnk}
% \usepackage{url}
\def\UrlBreaks{\do\/\do-}
% \usepackage{breakurl}
% \usepackage[breaklinks]{hyperref}
\usepackage{booktabs}

\usepackage{xspace,xpunctuate}
\newcommand{\ie}{{i.e.,}\xspace}
\newcommand{\eg}{{e.g.,}\xspace}
\newcommand{\ea}{{et~al\xperiod}\xspace}
\newcommand{\aka}{{a.k.a.}\xspace}
\newcommand{\etc}{{etc\xperiod}\xspace}
\setlength{\extrarowheight}{0.16cm}

% end of prologue

% \input{EGauthorGuidelines-body.inc}

% ---------------------------------------------------------------------
% EG author guidelines plus sample file for EG publication using LaTeX2e input
% D.Fellner, v2.03, Dec 14, 2018

\title[Chartability]%
      {How accessible is my visualization? Evaluating visualization accessibility with Chartability}

% for anonymous conference submission please enter your SUBMISSION ID
% instead of the author's name (and leave the affiliation blank) !!
% for final version: please provide your *own* ORCID in the brackets following \orcid; see https://orcid.org/ for more details.
% \author[Submission 1160]{Submission ID 1160}

\author[F. Elavsky, C. Bennett, \& D. Moritz]
{\parbox{\textwidth}{\centering Frank Elavsky,$^{1}$\orcid{0000-0002-6849-5893}
        Cynthia Bennett,$^{1}$
        and Dominik Moritz$^{1}$\orcid{0000-0002-3110-1053}
%        S. Spencer$^2$\thanks{Chairman Siggraph Publications Board}  
        }
        \\
% For Computer Graphics Forum: Please use the abbreviation of your first name.
{\parbox{\textwidth}{\centering $^1$Carnegie Mellon University, Human-Computer Interaction Institute\\
%        $^2$ Another Department to illustrate the use in papers from authors
%             with different affiliations
       } 
}
}
% ------------------------------------------------------------------------

% if the Editors-in-Chief have given you the data, you may uncomment
% the following five lines and insert it here
%
% \volume{36}   % the volume in which the issue will be published;
% \issue{1}     % the issue number of the publication
% \pStartPage{1}      % set starting page

%-------------------------------------------------------------------------
\begin{document}

\tabcolsep = 7.0pt
\def\arraystretch{0.8}

% \teaser{
%  \includegraphics[width=\linewidth]{eg_new}
%  \centering
%   \caption{New EG Logo}
% \label{fig:teaser}
% }

\maketitle
%-------------------------------------------------------------------------
\begin{abstract}
Novices and experts have struggled to evaluate the accessibility of data visualizations because there are no common shared guidelines across environments, platforms, and contexts in which data visualizations are authored. Between non-specific standards bodies like WCAG, emerging research, and guidelines from specific communities of practice, it is hard to organize knowledge on how to evaluate accessible data visualizations. We present Chartability, a set of heuristics synthesized from these various sources which enables designers, developers, researchers, and auditors to evaluate data-driven visualizations and interfaces for visual, motor, vestibular, neurological, and cognitive accessibility. In this paper, we outline our process of making a set of heuristics and accessibility principles for Chartability and highlight key features in the auditing process. Working with participants on real projects, we found that data practitioners with a novice level of accessibility skills were more confident and found auditing to be easier after using Chartability. Expert accessibility practitioners were eager to integrate Chartability into their own work. Reflecting on Chartability’s development and the preliminary user evaluation, we discuss tradeoffs of open projects, working with high-risk evaluations like auditing projects in the wild, and challenge future research projects at the intersection of visualization and accessibility to consider the broad intersections of disabilities.

%-------------------------------------------------------------------------
%  ACM CCS 1998
%  (see https://www.acm.org/publications/computing-classification-system/1998)
% \begin{classification} % according to https://www.acm.org/publications/computing-classification-system/1998
% \CCScat{Computer Graphics}{I.3.3}{Picture/Image Generation}{Line and curve generation}
% \end{classification}
%-------------------------------------------------------------------------
%  ACM CCS 2012
%   (see https://www.acm.org/publications/class-2012)
%The tool at \url{http://dl.acm.org/ccs.cfm} can be used to generate
% CCS codes.
%Example:
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003120.10003145.10011770</concept_id>
<concept_desc>Human-centered computing~Visualization design and evaluation methods</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003120.10011738.10011774</concept_id>
<concept_desc>Human-centered computing~Accessibility design and evaluation methods</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003120.10003121.10003122.10010855</concept_id>
<concept_desc>Human-centered computing~Heuristic evaluations</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Human-centered computing~Visualization design and evaluation methods}
\ccsdesc[500]{Human-centered computing~Accessibility design and evaluation methods}
\ccsdesc[500]{Human-centered computing~Heuristic evaluations}

\printccsdesc   
\end{abstract}
%-------------------------------------------------------------------------
\section{Introduction}

26\% of people in the United States self-report living with at least one disability~\cite{cdc_disability_2018}. Of those, 13.7\% live with a mobility disability and 10.8\% with a cognitive disability. Globally, the World Health Organization reports that 29\% of the world lives with uncorrected or uncorrectable blindness, low vision, or moderate to severe visual impairment~\cite{noauthor_world_nodate}. Access is a significant inclusion effort that has broad international impact, especially for data visualization.

Accessibility is the practice of making information, content, and functionality fully available to and usable by people with disabilities. As part of this process, practitioners need to be able to identify accessibility barriers. While general accessibility standards help, evaluating the inaccessibility of complex data systems can be a daunting and often expensive task. State-of-the art automated compliance checkers only find 57\% of accessibility errors~\cite{noauthor_study_2021}, meaning accessible experiences must still be manually designed and checked for quality. And following standards may only account for up to half of the needs of people with disabilities \cite{power_2012} anyway. Additionally, the intended wide applicability of these general standards means they fall short for information-rich systems, such as data visualizations (which use size, color, angles, shapes, and other dimensions to encode information). These specific contexts, communities, and libraries that deal with data visualizations and information-rich interfaces often have their own tools and guidelines for use, but they seldom include accessibility. Finally, research at the intersection of data visualization and accessibility has yet to meaningfully permeate data visualization tools and communities and primarily focuses on blindness and low vision, neglecting diverse accessibility needs of people with other disabilities.

Synthesizing evolving accessibility standards, research findings, and artifacts from communities of practice into usable knowledge for a specific, evolving domain is a wicked problem. To address this, we present Chartability. Chartability is an accessibility evaluation system specific to data visualizations and interfaces which aims to help practitioners answer the question, ``how accessible is my data visualization?'' Chartability organizes knowledge from disparate bodies of work into testable heuristics based on the functional accessibility principles POUR (Perceivable, Operable, Understandable, and Robust)~\cite{initiative_wai_accessibility_nodate} and 3 novel principles CAF (Compromising, Assistive, and Flexible), which we added to attend to the unique qualities and demands of data visualizations. We refer to these 7 heuristic principles as POUR+CAF. Chartability is a community-contributed project that leverages the governance strategies of open source projects as a way to address the complex dual-evolution of both accessibility and data interaction practices.

We additionally present an initial, light evaluation of Chartability from the experience of practitioners using it. We set out to see if using Chartability reduces the barrier of entry into this work for accessibility novices and if accessibility experts had any feedback to share about its use. We gave practitioners introductory material for Chartability and instructed them to use it according to their needs. We found that before using Chartability only accessibility experts believed auditing data visualizations to be somewhat easy or easy, while the other group believed auditing data visualizations to be somewhat hard or hard. All novice accessibility practitioners became more confident after using Chartability and believed auditing data visualizations for accessibility to be less difficult. Conversely while the expert accessibility practitioners were already confident in their ability to evaluate accessibility (and all unanimously had no change in their before and after evaluations), they were excited to adopt Chartability into their set of auditing resources.

Our work sets out to acknowledge that data practitioners face significant barriers when first making data visualizations, systems, and experiences accessible. While Chartability contributes to filling gaps and organizing knowledge, it also challenges visualization and data interaction researchers to explore new horizons of possibilities in this space. As such, we conclude with recommendations for future research at the crossroads of data visualization and accessibility.

\section{Existing Work in Data Visualization and Accessibility}

While recent works at the intersection of data visualization and accessibility are promising, they do not provide a consistent and unified methodology for designers to evaluate the accessibility of their work across the broad spectrum of disability considerations.

\subsection{Research Advancements in Data Visualization and Accessibility}
In parallel to Mack \ea's ``What do we mean by Accessibility Research?''~\cite{mack_what_2021} when we asked ``What do we mean by data visualization accessibility research?'' we found that nearly all topics of study were vision-related. Largely, access issues other than vision that affect data visualization (such as cognitive/neurological, vestibular, and motor concerns) are almost entirely unserved in this research space. Kim \ea found that 56 papers have been published between 1999 and 2020 that focus on vision-related accessibility (not including color vision deficiency), with only 3 being published at a visualization venue (and only recently since 2018)~\cite{kim_accessible_2021}. Marriott \ea found that there is no research at all that engages motor accessibility~\cite{marriott_inclusive_2021}. We have found 2 papers that engage cognitive/neurological disability in visualization and 1 student poster from IEEE Vis, which are all recent (specifically intellectual developmental disabilities~\cite{wu_understanding_2021} and seizure risk~\cite{south_generating_2020,south_detecting_2021}). We found no papers that engage vestibular accessibility, such as motion and animation-related accessibility. We also found that there is no research specific to low vision disabilities (not blindness or color vision deficiency) unless conflated with screen reader usage in data visualization. Blind and low vision people are often researched together, but in practice may use different assistive technologies (such as magnifiers and contrast enhancers) and have different interaction practices (such as a combination of sight, magnification, and screen reader use)~\cite{szpiro_2016}.

Since the 1990s, the most prominent and active accessibility topic in visualization has been color vision deficiency~\cite{chaparro_applications_2017,nunez_optimizing_2018,oliveira_towards_2013,9023497,martinez_methodology_2021}. Research projects that explore tactile sensory substitutions have been a topic in computational sciences dating back to the 1983~\cite{geldard_tactual_1983}, with tactile sensory substitutions being used for maps and charts as far back as the 1830s~\cite{noauthor_extensive_2016}. Sonification used both in comparison to and alongside visualization and tactile methods for accessibility dates as far back as 1985~\cite{mansur_sound_1985,flowers_cross-modal_1997,brewster_visualization_2002,mcgookin_soundbar_2006,zhao_data_2008,cullen_co-designing_2019}. Some more recent work has explored robust screen reader data interaction techniques~\cite{miesenberger_accessible_2018,sorge_polyfilling_2016}, screen reader user experiences with digital, 2-D spatial representations, including data visualizations~\cite{schaadhardt_understanding_2021,sharif_understanding_2021}, dug deeper into the semantic layers of effective chart descriptions~\cite{lundgard_accessible_22}, and investigated how to better understand the role of sensory substitution~\cite{chundury_towards_2022}. Jung \ea offer guidance that expands beyond commonly cited literature that chart descriptions are preferably between 2 and 8 sentences long, written in plain language, and with consideration for the order of information and navigation~\cite{jung_communicating_2022}. We find all of this emerging work promising and foundational.

Despite this promising work emerging, we also want to acknowledge a spectrum of other work that exists at the intersection of accessibility and data visualization that does not serve the goals of our project. There is significant research that explores automatic or extracted textual descriptions~\cite{choi_visualizing_2019,balaji_chart-text_2018,chen_neural_2019,chen_figure_2020,lai_automatic_2020,obeid_chart--text_2020,qian_generating_2021,sharif_evographs_2018} and haptic graphs and tactile interfaces~\cite{aldrich_talk_2008,bornschein_collaborative_2015, brown_viztouch_2012,butler_technology_2021,gallace_what_2011,geldard_tactual_1983,jansen_opportunities_2015,jayant_automated_2007,lederman_perception_1986,schneider_constructing_nodate,shi_tickers_2016,baker_2016}. These research projects produce artifacts that are high-cost for individual use, some are not robust enough to interpret complex visualizations effectively, and several have not included people with disabilities. Since our goal is to synthesize knowledge for practitioner accessibility work, we also acknowledge that some of these projects did not follow standards during their research project and in their output, such as using Web Content Accessibility Guidelines~\cite{noauthor_web_nodate} or The American Printing House for the Blind and Braille Authority of North America~\cite{noauthor_guidelines_nodate}. All of these challenges are factors that limit the generalizability of these artifacts and knowledge for practitioner use~\cite{lundgard_sociotechnical_2019,moraes_evaluating_2014,sharif_understanding_2021}. We encourage work to continue at the intersection of accessibility and visualization, but stress the importance of practical, disability-led research that either builds on or explicitly challenges standards.

\subsection{Accessibility Practices in Data Visualization Tools and Libraries}
Our research goals are to find what is already being done in data visualization and accessibility and to see if we can enhance that activity. To this aim, our background investigation includes a broad and comprehensive exploration of the field of practitioner and non-academic artifacts.

Some open source and industry contributions have pushed data visualization and related accessibility efforts. Libraries like Highcharts~\cite{noauthor_accessibility_nodate} or Visa Chart Components (VCC)~\cite{vcc} and tools like the Graphics Accelerator in SAS~\cite{noauthor_sas_nodate} have broad accessibility functionality built in, but their documentation is technically specific to their implementation. While these relatively accessible libraries and tools can be helpful for inspiration, their specific techniques and guidance materials are not easily transferrable to other environments or applications where data visualizations are created. Practitioners must reverse engineer and deconstruct many of the methods employed by these libraries, and with the exception of VCC (which is open source), this task requires significant effort, given their primarily closed-box nature.

In common charting tools and libraries (apart from those already mentioned) accessibility engineering is often not present, limited in scope, or has only recently become an effort. More established visualization libraries like matplotlib, ggplot2, d3js, R-Shiny, and Plotly have left most accessibility efforts to developers, with varying levels of documentation and difficulty involved~\cite{noauthor_are_2018, noauthor_making_2018, noauthor_revealing_nodate, noauthor_solved_2019, simon_making_2020}. None of these major tools have a broad spectrum of accessibility options built in and documented.

>>>>

\documentclass[journal]{vgtc}                % final (journal style)
% \documentclass[review,journal]{vgtc}         % review (journal style)
%\documentclass[widereview]{vgtc}             % wide-spaced review
%\documentclass[preprint,journal]{vgtc}       % preprint (journal style)

%% Uncomment one of the lines above depending on where your paper is
%% in the conference process. ``review'' and ``widereview'' are for review
%% submission, ``preprint'' is for pre-publication, and the final version
%% doesn't use a specific qualifier.

%% Please use one of the ``review'' options in combination with the
%% assigned online id (see below) ONLY if your paper uses a double blind
%% review process. Some conferences, like IEEE Vis and InfoVis, have NOT
%% in the past.

%% Please use the ``preprint''  option when producing a preprint version
%% for sharing your article on an open access repository

%% Please note that the use of figures other than the optional teaser is not permitted on the first page
%% of the journal version.  Figures should begin on the second page and be
%% in CMYK or Grey scale format, otherwise, colour shifting may occur
%% during the printing process.  Papers submitted with figures other than the optional teaser on the
%% first page will be refused. Also, the teaser figure should only have the
%% width of the abstract as the template enforces it.

%% These few lines make a distinction between latex and pdflatex calls and they
%% bring in essential packages for graphics and font handling.
%% Note that due to the \DeclareGraphicsExtensions{} call it is no longer necessary
%% to provide the path and extension of a graphics file:
%% \includegraphics{diamondrule} is completely sufficient.
%%
\ifpdf%                                % if we use pdflatex
  \pdfoutput=1\relax                   % create PDFs from pdfLaTeX
  \pdfcompresslevel=9                  % PDF Compression
  \pdfoptionpdfminorversion=7          % create PDF 1.7
  \ExecuteOptions{pdftex}
  \usepackage{graphicx}                % allow us to embed graphics files
  \DeclareGraphicsExtensions{.pdf,.png,.jpg,.jpeg} % for pdflatex we expect .pdf, .png, or .jpg files
\else%                                 % else we use pure latex
  \ExecuteOptions{dvips}
  \usepackage{graphicx}                % allow us to embed graphics files
  \DeclareGraphicsExtensions{.eps}     % for pure latex we expect eps files
\fi%

%% it is recomended to use ``\autoref{sec:bla}'' instead of ``Fig.~\ref{sec:bla}''
\graphicspath{{figures/}{pictures/}{images/}{./}} % where to search for the images

\usepackage{microtype}                 % use micro-typography (slightly more compact, better to read)
\PassOptionsToPackage{warn}{textcomp}  % to address font issues with \textrightarrow
\usepackage{textcomp}                  % use better special symbols
\usepackage{mathptmx}                  % use matching math font
\usepackage{times}                     % we use Times as the main font
\renewcommand*\ttdefault{txtt}         % a nicer typewriter font
\usepackage{cite}                      % needed to automatically sort the references
\usepackage{tabu}                      % only used for the table example
\usepackage{booktabs}                  % only used for the table example
\usepackage{listings}
\usepackage[skip=3pt]{caption}
%% We encourage the use of mathptmx for consistent usage of times font
%% throughout the proceedings. However, if you encounter conflicts
%% with other math-related packages, you may want to disable it.

\usepackage{xspace,xpunctuate}
\newcommand{\ie}{{i.e.,}\xspace}
\newcommand{\eg}{{e.g.,}\xspace}
\newcommand{\ea}{{et~al\xperiod}\xspace}
\newcommand{\aka}{{a.k.a.}\xspace}
\newcommand{\etc}{{etc\xperiod}\xspace}

%% In preprint mode you may define your own headline. If not, the default IEEE copyright message will appear in preprint mode.
%\preprinttext{To appear in IEEE Transactions on Visualization and Computer Graphics.}

%% In preprint mode, this adds a link to the version of the paper on IEEEXplore
%% Uncomment this line when you produce a preprint version of the article 
%% after the article receives a DOI for the paper from IEEE
%\ieeedoi{xx.xxxx/TVCG.201x.xxxxxxx}

%% If you are submitting a paper to a conference for review with a double
%% blind reviewing process, please replace the value ``0'' below with your
%% OnlineID. Otherwise, you may safely leave it at ``0''.
\onlineid{1580}

%% declare the category of your paper, only shown in review mode
\vgtccategory{Research}
%% please declare the paper type of your paper to help reviewers, only shown in review mode
%% choices:
%% * algorithm/technique
%% * application/design study
%% * evaluation
%% * system
%% * theory/model
\vgtcpapertype{Systems \& Rendering}

%% Paper title.
\title{Data Navigator: An Accessibility-Centered Data Navigation Toolkit}

%% This is how authors are specified in the journal style

%% indicate IEEE Member or Student Member in form indicated below
% \author{Anonymous}
\author{Frank Elavsky, Lucas Nadolskis, Dominik Moritz}
\authorfooter{
%% insert punctuation at end of each item
\item
 All authors are with Carnegie Mellon University.
 \item
 Frank Elavsky: fje@cmu.edu.
 \item
 Lucas Nadolskis: nadolskis@cmu.edu.
\item
 Dominik Moritz: domoritz@cmu.edu.
}

%other entries to be set up for journal
\shortauthortitle{Elavsky: Data Navigator}
% \shortauthortitle{Elavsky \MakeLowercase{\textit{et al.}}: Data Navigator}
%\shortauthortitle{Firstauthor \MakeLowercase{\textit{et al.}}: Paper Title}

%% Abstract section.
% The lack of accessible data visualizations for people with disabilities remains a significant challenge in current toolkit and practitioner efforts. Existing visualizations often lack underlying structure, fail to engage necessary input modalities, and rely heavily on visual rendering practices. This limitation stems from a scarcity of building materials, such as non-pointer data navigation structures and accessible rendering formats beyond scalable vector graphics (SVG).

% To address this gap, we present Data Navigator, a toolkit based on a graph data structure that supports various data structures and input modalities. It provides an interface for screen reader, keyboard, touch gestures, motion gestures, voice, and custom input methods. By decoupling navigation and visual encodings, Data Navigator empowers visualization practitioners with expressive control over their designs. We offer a high-level system design and a web-based implementation using JavaScript, HTML, and CSS. Through case examples, we demonstrate the replication of best practices, integration with existing toolkits, and development of novel prototypes for accessible navigation.

% Data Navigator represents a step towards making more accessible data experiences easier to design and implement, involving both toolkit makers and visualization practitioners.
\abstract{Making data visualizations accessible for people with disabilities remains a significant challenge in current practitioner efforts. Existing visualizations often lack an underlying navigable structure, fail to engage necessary input modalities, and rely heavily on visual-only rendering practices. These limitations exclude people with disabilities, especially users of assistive technologies. To address these challenges, we present Data Navigator: a system built on a dynamic graph structure, enabling developers to construct navigable lists, trees, graphs, and flows as well as spatial, diagrammatic, and geographic relations. Data Navigator supports a wide range of input modalities: screen reader, keyboard, speech, gesture detection, and even fabricated assistive devices. We present 3 case examples with Data Navigator, demonstrating we can provide accessible navigation structures on top of raster images, integrate with existing toolkits at scale, and rapidly develop novel prototypes. Data Navigator is a step towards making accessible data visualizations easier to design and implement.%
% There is a gap between techniques we know are effective for accessible data navigation and what tools are currently capable of. Although scalable vector graphics (SVG) and raster images are common formats for data visualizations, they fail to express relationships in a meaningful way. This has resulted in a reliance on tree or list structures inherent in many UI paradigms, which are not generic enough to express relational, spatial, diagrammatic, and geographic information common to visualization. To address these limitations, we propose Data Navigator, a toolkit built on a graph data structure.
} % end of abstract

%% Keywords that describe your work. Will show as 'Index Terms' in journal
%% please capitalize first letter and insert punctuation after last keyword
\keywords{accessibility, visualization, tools, technical materials, platforms, data interaction}

%% ACM Computing Classification System (CCS). 
%% See <http://www.acm.org/class/1998/> for details.
%% The ``\CCScat'' command takes four arguments.

% \begin{CCSXML}
% <ccs2012>
%    <concept>
%        <concept_id>10003120.10003145.10003151</concept_id>
%        <concept_desc>Human-centered computing~Visualization systems and tools</concept_desc>
%        <concept_significance>500</concept_significance>
%        </concept>
%    <concept>
%        <concept_id>10003120.10011738.10011776</concept_id>
%        <concept_desc>Human-centered computing~Accessibility systems and tools</concept_desc>
%        <concept_significance>500</concept_significance>
%        </concept>
%    <concept>
%        <concept_id>10003120.10003123.10011760</concept_id>
%        <concept_desc>Human-centered computing~Systems and tools for interaction design</concept_desc>
%        <concept_significance>300</concept_significance>
%        </concept>
%  </ccs2012>
% \end{CCSXML}

% \ccsdesc[500]{Human-centered computing~Visualization systems and tools}
% \ccsdesc[500]{Human-centered computing~Accessibility systems and tools}
% \ccsdesc[300]{Human-centered computing~Systems and tools for interaction design}

%% A teaser figure can be included as follows

\teaser{
  \includegraphics[width=\textwidth]{data_navigator.png}
  \caption{Data Navigator provides data visualization libraries and toolkits with accessible data navigation structures, robust input handling, and flexible semantic rendering capabilities.}
  \label{fig:teaser}
}

%% Uncomment below to disable the manuscript note
%\renewcommand{\manuscriptnotetxt}{}

%% Copyright space is enabled by default as required by guidelines.
%% It is disabled by the 'review' option or via the following command:
% \nocopyrightspace

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%% START OF THE PAPER %%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

%% The ``\maketitle'' command must be the first command after the

%% ``\begin{document}'' command. It prepares and prints the title block.

%% the only exception to this rule is the \firstsection command

\firstsection{Introduction}

\maketitle

While there is a growing interest in making data visualizations more accessible for people with disabilities, current toolkit and practitioner efforts have not risen to the challenge at scale. Major data visualization tools and ecosystems predominantly produce inaccessible artifacts for many users with disabilities. We believe this is largely a gap caused by a lack of underlying structure in most visualizations, failure to engage the input modalities used by people with disabilities, and over-reliance on visual-only rendering practices.

Users who are blind or low vision commonly use screen readers and users with motor and dexterity disabilities often do not use "pointer" (precise mouse and touch) based input technology when interacting with digital interfaces. Many users with motor and dexterity disabilities use discrete navigation controls, either sequentially using keyboard-like input, or directly using voice or text commands.

Most interactive visualizations simply focus on pointer-based input: they can be clicked or tapped, hovered, and selected in order to perform analytical tasks. This excludes non-pointer input technologies. These devices require consideration for the navigation structure and underlying semantics of a visual interface.

However, building navigable spatial and relational interfaces is a difficult task with current resources.

Raster images, arguably the most common format for creating and disseminating data visualizations, currently cannot be made into navigable structures. These are only described using alt text, which limits their usefulness to screen reader users.

Unfortunately, more accessible rendering formats like SVG with ARIA (accessible rich internet applications) properties are more resource intensive than raster approaches, like WebGL-powered HTML canvas or pre-rendered PNG files. SVG puts a burden on low-bandwidth users and a ceiling on how many data points can be rendered in memory.

In addition, ARIA itself has 2 major limitations. First, when added to interface elements, ARIA only provides \textit{screen reader} access, which means that developers must build a solution from scratch for other navigation input modalities. Second, ARIA's linear navigation structure can be time-consuming for screen reader users if a visualization has many elements. This may impede how essential insights and relationships are understood~\cite{Sorge2016Polyfilling,Godfrey2018Accessible,Zong2022Rich,Thompson2023Chart,Jung2022Communicating,Sharif2021Understanding}.

Some emerging approaches have sought to address this serial limitation of data navigation and provide richer experiences for screen reader users~\cite{Sorge2016Polyfilling,Godfrey2018Accessible,Zong2022Rich,Thompson2023Chart}. However, these approaches rely on a tree-based navigation structure which is often not an appropriate choice for visualizations of relational, spatial, diagrammatic, or geographic data. Many visualization structures are currently unaddressed.

Zong \ea stress that in order to realize richer, more accessible data visualizations, the responsibility must be shared by ``toolkit makers,'' the practitioners who design, build, and maintain visualization authoring technologies~\cite{Zong2022Rich}. Our contribution is towards that aim, to make more accessible data experiences easier to design and implement within existing visualization work.

We present Data Navigator. Data Navigator is a toolkit built on a graph data structure, within which a broad array of common data structures can be expressed (including list, tree, graph, relational, spatial, diagrammatic, and geographic structures). Data Navigator also exposes an interface that supports interactions via screen reader, keyboard, gesture-based touch, motion gesture, voice, as well as fabricated and DIY input modalities. Data Navigator provides expressive structure and semantic rendering capabilities as well as the ability for developers to use their own, preferred method of rendering.

Data Navigator builds upon human-studies motivated work on accessible navigation~\cite{Zong2022Rich,Thompson2023Chart} towards a more generalizable resource for visualization practitioners. We contribute a high-level system design for our node-edge graph-based solution as well as an implementation of this system on the web, using JavaScript, HTML, and CSS. Through our case examples we also demonstrate that our generalized approach is suitable for replication of existing best practices from other systems, integration into existing visualization toolkit ecosystems, and development of novel prototypes for accessible navigation. We illustrate how Data Navigator's use of generic edges, dynamic navigation rules, and loose coupling between navigation and visual encodings provides practitioners robust, expressive, control over their system designs.

\section{Related Work}
Our contribution is an attempt to bridge the gap between research and practice more effectively across broad ecosystems in order to enable deeper and more expressive accessible data navigation interfaces. Below we outline the prior research and standards that inform our project, a breakdown of existing visualization toolkit approaches to data navigation, and then accessible input device considerations.

\subsection{Accessibility research and standards in visualization}
Research and standards are both somewhat limited by a strong bias towards visual disabilities. In \textit{Chartability}, 36 of the 50 criteria related to accessible visualization considerations involve visual disabilities~\cite{Elavsky2022Chartability, Fan2023Accessibility}. Marriott \ea also found that visual disability considerations are the primary focus of data visualization literature~\cite{Marriott2021Inclusive}, leaving the barriers that many other demographics face unstudied.

However, despite the heavy focus on visual disabilities, the work that does exist in the visualization community is deeply valuable and serves as an important starting point for our technical contribution.

\subsubsection{Accessible navigation design considerations}

Zong \ea's research, which was conducted as in-depth co-design work and validated in usability studies involving blind participants, presented a design space for accessible, rich screen reader navigation of data visualizations. They organized their design space into \textit{structure, navigation,} and \textit{description} considerations and demonstrated example \textit{structural}, \textit{spatial}, and \textit{direct} tree-based approaches~\cite{Zong2022Rich}.

\textit{Chart Reader} also engaged these design space considerations in their co-design work on accessible data navigation structures~\cite{Thompson2023Chart}. We consider these design dimensions as the best starting point for our work, bridging the gap between research and toolkits.

There are additional research projects that have focused on accessible data navigation and interaction ~\cite{Sharif2022VoxLens,Sorge2016Polyfilling,Godfrey2018Accessible,Sharif2018evoGraphs}. These contributions explore a range of different interaction structures, including lists, trees, and tables of information as well as direct access methods such as voice interface commands and simple, pre-determined questions.

\subsubsection{Accessible visualization: understanding users}

A wide array of emerging research projects investigate screen reader users needs, barriers, and preferences, and offer guidelines, models, and considerations for creating accessible data visualizations~\cite{Sharif2021Understanding,Lundgard2022Accessible,Chundury2022Towards,Fan2023Accessibility}. Jung \ea offer guidance to consider the order of information in textual descriptions and during navigation~\cite{Jung2022Communicating}. Kim \ea collected screen reader users' questions when interacting with data visualizations, which could open the door for more natural language data interaction~\cite{Kim2023Exploring}.

\subsubsection{Accessibility standards and guidelines}
In the space of research, there has been a growing interest in developing guidelines for practitioners~\cite{Elavsky2022Chartability, Durant2022Ten} and even applying guidelines as a method of validation alongside human studies evaluations and co-design~\cite{Lundgard2019Sociotechnical,Zong2022Rich,Lundgard2022Accessible,Fan2023Accessibility}. Unfortunately, most accessibility standards and guidelines do not explicitly engage how to structure data navigation.

Despite this, existing accessibility standards bodies like the Web Content Accessibility Guidelines do stress the importance of accurate, functional semantics in order for screen reader users to know how to interact with elements~\cite{WAI2021Semantics}. For interactive visualizations this means that button-like or link-like behavior should expressly be made using elements that are semantically buttons and links. Our system should be capable of expressing meaningful semantics to users of assistive technologies.

\subsection{Visualization toolkits and technical work}
Unfortunately while many data visualization toolkits offer some degree of accessible navigation and interaction capabilities to developers, very few toolkits currently out there offer control over the important aspects of accessible data navigation design. Replicating existing research and strategies, remediating toolkit ecosystems, and building novel prototypes are all difficult or impossible to do due to the current lack of toolkit capabilities.

Existing data visualization toolkits have 3 major limitations that we wanted to address in the design of Data Navigator:
\begin{enumerate}
  \itemsep-0.4em
  \item \textbf{Built on visual materials}: toolkits produce either raster or SVG-based visualizations, neither of which are focused towards designing navigable, semantic structures. As a consequence, many visualizations are simply entirely inaccessible.
  \item \textbf{Lacking relational expressiveness}: When data navigation \textit{is} provided, the navigation is based on either a tree or list structure (see~\autoref{existing}). The consequence of this limitation is that many other non-list and non-tree data relationships become difficult or impossible to represent without overly tedious navigation or inefficient architecture.
  \item \textbf{Designed only for screen reader interaction}: When \textit{accessible} data navigation is provided, it is generally only made possible through SVG with ARIA (Accessible Rich Internet Application) attributes. ARIA is primarily only leveraged by screen readers~\cite{WAI2017ARIA}. If a data element can be clicked and performs some form of function, only direct pointer (mouse and touch) and screen reader users are included. The consequence of this is that a wide array of other input devices, many used as assistive technologies by people with motor and dexterity disabilities, are excluded.
\end{enumerate}

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{figures/state-of-the-art.png}
  \caption{Existing accessibility trees and lists, shown using node-edge graph conventions. (*) Denotes only \textit{screen reader} access. (**) Denotes \textit{screen reader}, \textit{keyboard-only}, and \textit{pointer} access as well.}
  % \Description{A mouse pointer is shown shaking over many data points that are very small while a touch hit area is shown over the same visualization, overlapping with many points at once.}
  \label{existing}
\end{figure}
